![](Misc/Gen_Re_logo.svg){width="335"}

# Workshop Title: Data Analytics in R

This workshop will cover some of the steps we need to sanitise the data before we can do analysis in R. This will be a hands-on session.

The key focus areas for this session are:

-   Setting up an R Project

-   Libraries that are commonly used

-   Data Loading

-   Data Exploration

-   Data Transformation

-   Analysis using IBNR calculation

Take-aways are:

-   A high-level understanding of what R can do.

-   How to properly set up a project for use.

-   Data preparation and manipulation in R.

-   Analysis with R.

This session will use the pseudo data provided by us. Our goal is to \_\_\_\_\_\_\_\_\_.

## 1) Loading and Installing Libraries

#### Method 1

```{r , warning=FALSE, message=FALSE}

# Read Packages
library(readxl)       # Read xlsx format (fast)
library(writexl)      # Write xlsc format (fast)
library(haven)        # Import and Export SPSS, Stata and SAS Files

# Data Summary Packages
library(skimr)
library(DataExplorer)

# Data Transformation Packages
library(data.table) 
library(dplyr)

# Tool Packages
library(lubridate)
library(stringr)
library(zoo)
library(janitor)

# Multitread Packages
library(doParallel)
library(foreach)


```

#### Method 2

```{r}

# source("Dependecies.R")
```

## 2) Setting Display and Scientific Penalties

```{r}

rm(list=ls()) # Removes all objects in R environment

gc() # This clears the cache to prep. Automatically releases memory when an object is no longer used. It does this by tracking how many names point to each object, and when there are no names pointing to an object, it deletes that object.



options(scipen=999)       # Maximum accuracy
options(max.print = 9999) # Printing Options

dir.create("RData/0. Misc Data") # Creating Folder for collection of Data Errors
End.Suvery <- ymd("20200101") # Create End Point for Study
```

#### Reading in Data

```{r}
Policy_data <- read.csv("Files/Policy Data.csv") # Savepoint / Checkpoint
Claims_data <- read.csv("Files/Claims Data.csv") # Savepoint / Checkpoint

# Set as Data Table format
setDT(Policy_data)
setDT(Claims_data)
# 
# saveRDS(Policy_data, "RData/0. Raw Data/Policy_Data.rds")
# saveRDS(Claims_data, "RData/0. Raw Data/Claims_Data.rds")
# Policy_data <- readRDS("RData/0. Raw Data/Policy_Data.rds")
# Policy_data <- readRDS("RData/0. Raw Data/Policy_Data.rds")

```

## 3) Mapping / Renaming Data

#### Method 1

```{r}

# Changing the names for the column into a unified format and without spacing

setnames( Policy_data, "Policy.Number"                        , "i_polnum")
setnames( Policy_data, "Plan.Code..Dread.Disease."            , "i_plancd_dd")
setnames( Policy_data, "Plan.Code..Basic.Plan."               , "i_plancd_ba")
setnames( Policy_data, "Effective.Date"                       , "i_effdt")
setnames( Policy_data, "Sales.Channel"                        , "i_chan")
setnames( Policy_data, "Benefit.Term"                         , "i_bentm")
setnames( Policy_data, "Premium.Payment.Term"                 , "i_prmtm")
setnames( Policy_data, "End.of.Survey.Period.Coverage.Status" , "i_endsts")
setnames( Policy_data, "Termination.Date"                     , "i_terdt")
setnames( Policy_data, "Face.Amount..Dread.Disease.Benefit."  , "i_face_dd")
setnames( Policy_data, "Insured.ID"                           , "i_insdid")
setnames( Policy_data, "Date.of.Birth"                        , "i_dob")
setnames( Policy_data, "Sex"                                  , "i_sex")


setnames( Claims_data, "Policy.Number"                        ,"c_polnum")
setnames( Claims_data, "Plan.Code..Dread.Disease."            ,"c_plancd_dd")
setnames( Claims_data, "Plan.Code..Basic.Plan."               ,"c_plancd_ba")
setnames( Claims_data, "Effective.Date"                       ,"c_effdt")
setnames( Claims_data, "Event.Date"                           ,"c_occdt")
setnames( Claims_data, "Face.Amount..Dread.Disease.Benefit."  ,"c_face_dd")
setnames( Claims_data, "Insured.ID"                           ,"c_insdid")
setnames( Claims_data, "Date.of.Birth"                        ,"c_dob")
setnames( Claims_data, "Sex"                                  ,"c_sex")
setnames( Claims_data, "Claim.amount"                         ,"c_clmamt")
setnames( Claims_data, "Notification.Date"                     ,"c_notdt")
setnames( Claims_data, "Settlement.Date"                       ,"c_setldt")
setnames( Claims_data, "Claim.Status"                          ,"c_clmsts")
setnames( Claims_data, "Claim.Category"                        ,"c_clmct")
setnames( Claims_data, "Cause.Name"                            ,"c_cause_name")

```

#### Method 2

```{r, eval=FALSE}

# Here, we create and external data table that we can use to rename all together which is quicker and shorter
## For more advanced users: we can create an excel to store the column names. Then type the ideal column names and read it back in

# col_name_mapping <- 
#   data.table( old_col = c("Policy.Number"                        ,
#                           "Plan.Code..Dread.Disease."            ,
#                           "Plan.Code..Basic.Plan."               ,
#                           "Effective.Date"                       ,
#                           "Sales.Channel"                        ,
#                           "Benefit.Term"                         ,
#                           "Premium.Payment.Term"                 ,
#                           "End.of.Survey.Period.Coverage.Status" ,
#                           "Termination.Date"                     ,
#                           "Face.Amount..Dread.Disease.Benefit."  ,
#                           "Insured.ID"                           ,
#                           "Date.of.Birth"                        ,
#                           "Sex"                                  ),
#               new_col = c( "i_polnum",
#                            "i_plancd_dd",
#                            "i_plancd_ba",
#                            "i_effdt",
#                            "i_chan",
#                            "i_bentm",
#                            "i_prmtm",
#                            "i_endsts",
#                            "i_terdt",
#                            "i_face_dd",
#                            "i_insdid",
#                            "i_dob",
#                            "i_sex"))
# 
# setnames(Policy_data, col_name_mapping$old_col, col_name_mapping$new_col )

```

## 4) EDA

#### Generating seq IDs for easier identification

```{r}

# Using 1:nrow(Policy_data) creates unique number for each row of dats - this will allow us to track the data easier
Policy_data[ ,`:=`(i_seqid = 1:nrow(Policy_data))]
Claims_data[ ,`:=`(c_seqid = 1:nrow(Claims_data))]
```

#### Summary of Data / Overview

```{r}

# Survey.Date.Start <- ymd("20150101")
# Survey.Date.End   <- ymd("20200101")

summary(Policy_data) # Base Function from R to get idea of the data
summary(Claims_data)

skimr::skim(Policy_data) # Skimming the data using external libraries to get better idea of the data
skimr::skim(Claims_data)

Policy_data[  , .N , .(is.na(i_polnum))] # Checking for missing policy number
Policy_data[ , .N ,.(i_plancd_dd)]       # Checking for count of each plancodes and if there are any missing
Policy_data[ , .N ,.(i_plancd_ba)]

```

#### Mapping of Product IDs from Plan Codes

```{r}

# Method 1 of mapping Plancode to Prdid - Using on Policy
Policy_data[ str_detect(i_plancd_dd, "A") , i_prdid := "A"]
Policy_data[ str_detect(i_plancd_dd, "B") , i_prdid := "B"]
Policy_data[ str_detect(i_plancd_dd, "C") , i_prdid := "C"]
Policy_data[ str_detect(i_plancd_dd, "Z") , i_prdid := "Z"]

Policy_data[ , table(i_plancd_dd, i_prdid, useNA = "ifany")]

# Method 2 of mapping Plancode to Prdid - Using on Claims
plancode_mapping <- 
  data.table( plncd  = c("A1","A2","A3","B1","B2","B3","C1","C2","C3","Z1","Z2","Z3"), prdid = c("A","A","A","B","B","B","C","C","C","Z","Z","Z"))

Claims_data <- merge(Claims_data, plancode_mapping, 
                     by.x = "c_plancd_dd",
                     by.y = "plncd",
                     all.x = T)
setnames(Claims_data, "prdid" , "c_prdid")
```

#### Mapping of Channel

```{r}
Policy_data[ , .N , .(i_chan)] # Check Channel Distribution

Policy_data[ i_chan  == "agency"                , i_chan := "A"]
Policy_data[ i_chan  == "broker"               , i_chan := "O"]
Policy_data[ i_chan  == "bank"                 , i_chan := "B"]
```

#### Finding Missing Values

##### Empty Data Row

```{r}

# Identified one line of empty data - to be removed
Policy_data[ is.na(i_polnum)]

saveRDS(Policy_data[ is.na(i_polnum)], "RData/0. Misc Data/Error Row.rds")

Policy_data <- Policy_data[ !is.na(i_polnum)]

```

##### Missing Gender

```{r}
Policy_data[ is.na(i_sex) , .N] # Checking for NA or missing gender
Policy_data[ i_sex == "" , .N]

Gender_table <- 
  Policy_data[ , table(i_insdid, i_sex, useNA = "ifany")] %>% 
  as.data.frame.matrix() %>%
  setDT( keep.rownames = T) # Create a gender table to detect double genders

Gender_table[ `F` > 0 & `M` > 0]

saveRDS(Policy_data[ i_insdid == 2067] , "RData/0. Misc Data/double_gender.rds")
Policy_data[ i_insdid == 2067,`:=`(i_sex = "F")]

Missing_gender <- Policy_data[ i_sex == ""]
saveRDS(Missing_gender,"RData/0. Misc Data/missing_gender.rds")
# Searching in Claims Data if there are possible claims 
Claims_data[ c_insdid %in% Missing_gender$i_insdid | 
               c_polnum %in% Missing_gender$i_polnum]


# Getting rough ditribution of gender to fill in missing data - can be fine tuned to product specific
Policy_data[ !(i_seqid %in% Missing_gender$i_seqid) & i_sex == "M" , .N ] / Policy_data[ !(i_seqid %in% Missing_gender$i_seqid) , .N]

Policy_data[ i_insdid %in% c("677", "4085", "909", "2795") , `:=`(i_sex  = "M")]
Policy_data[ i_insdid %in% c("4637", "789", "4329", "4854") , `:=`(i_sex  = "M")]

```

#### Date Checks

```{r}
Policy_data[ ,`:=`(i_effdt = mdy(i_effdt),
                   i_dob   = mdy(i_dob)  ,
                   i_terdt = mdy(i_terdt))] # Changing date formats to a standard date format

Policy_data[ i_terdt < i_dob   , .N] # count of Termination Date earlier than DOB
Policy_data[ i_terdt < i_effdt , .N] # count of Termination Date earlier than Effective Date
Policy_data[ i_effdt < i_dob   , .N] # count of Effective Date earlier than DOB

Policy_data[ i_effdt < i_dob, .(i_insdid, i_polnum, i_dob, i_effdt, i_terdt)]

Claims_data[ c_insdid == "904" | c_polnum == "140"]

saveRDS(Policy_data[ !(i_effdt < i_dob) ], "RData/0. Misc Data/effdt_after_dob.rds")

Policy_data <- Policy_data[ !(i_effdt < i_dob) ]

Claims_data[ is.na(c_occdt)  ]
Claims_data[ is.na(c_notdt)  ]
Claims_data[ is.na(c_setldt) ]
```

#### Mapping of End Status

```{r}
Policy_data[ , .N , .(i_endsts)]

# Renaming of values
Policy_data[ i_endsts  == "In-force"                , i_endsts := "I"] 
Policy_data[ i_endsts  == "Lapse"                   , i_endsts := "L"]
Policy_data[ i_endsts  == "Mature"                  , i_endsts := "M"]
Policy_data[ i_endsts  == "Terminated due to Claim" , i_endsts := "T"]

# Table to show Termination Date and End Status
#  - TRUE  : Termination Date is provided
#  - FALSE : Termination Date is no provided
Policy_data[ , table(i_endsts , !is.na(i_terdt) , useNA = "ifany")]

Policy_data[ is.na(i_terdt) & i_endsts == "L" ,.(i_insdid, i_effdt, i_bentm)]

# If Lapse and benefit term less than time period, we will lapse on survey end date
Policy_data[ is.na(i_terdt) & 
               i_endsts == "L" & 
               (year(i_effdt) + i_bentm) < 2019,
             i_terdt := End.Suvery]

# If Lapse and benefit term more than end date, we will keep it as in force
Policy_data[ is.na(i_terdt) & 
               i_endsts == "L" & 
               (year(i_effdt) + i_bentm) > 2019,
             i_endsts := "I" ]

# Policy_data[ is.na(i_terdt) & 
#                i_endsts == "L" & 
#                (year(i_effdt) + i_bentm) > 2019,
#              
#              `:=`(i_endsts = "S",
# s                  i_terdt  = End.Suvery) ]

```

#### 

#### Face Amount Checks

```{r}
Policy_data[ is.na(i_face_dd) | i_face_dd == 0]

Policy_data[ , .N , .(i_face_dd)][order(i_face_dd)]
Policy_data[ , .N , .(i_face_dd)][ , barplot(N,names.arg = sort(unique(na.omit(Policy_data$i_face_dd))), las=2 )] + 
  title(main = list("Policy - Barplot of DD Face Amount", font = 4)) 
  

Claims_data[ is.na(c_face_dd) | c_face_dd == 0]
Claims_data[ , .N , .(c_face_dd)][ , barplot(N,names.arg = sort(unique(na.omit(Claims_data$c_face_dd))), las=2 )] + 
  title(main = list("Barplot of DD Face Amount", font = 4)) 

Claims_data[ , .N , .(c_clmamt)][ , barplot(N,names.arg = sort(unique(na.omit(Claims_data$c_clmamt))), las=2 )] + 
  title(main = list("Claims - Barplot of Claim Amout", font = 4)) 

```

#### Benefit and Premium Term Check

```{r}
Policy_data[ ,.N , .(i_prmtm)]
Policy_data[ ,.N , .(i_prmtm)]
Policy_data[ i_prmtm > i_bentm , .N]
```

#### Duplicate Checks

```{r}
Policy_data[ ,`:=`(pseudocode_1 = paste(i_insdid, i_plancd_dd),
                   pseudocode_2 = paste(i_insdid, i_plancd_dd, i_dob))]

Policy_data[ duplicated(i_insdid)]
Policy_data[ duplicated(i_insdid) | duplicated(i_insdid, fromLast = T)]

Policy_data[ duplicated(pseudocode_1)]
Policy_data[ duplicated(pseudocode_1) | duplicated(pseudocode_1, fromLast = T)]
Policy_data[ duplicated(pseudocode_2)]
Policy_data[ duplicated(pseudocode_2) | duplicated(pseudocode_2, fromLast = T)]

saveRDS(Policy_data[ duplicated(pseudocode_1)], "RData/0. Misc Data/dup_records.rds")

Policy_data <- Policy_data[ !duplicated(pseudocode_2)]

Claims_data[ ,`:=`(pseudocode_1 = paste(c_insdid, c_plancd_dd),
                   pseudocode_2 = paste(c_insdid, c_plancd_dd, c_dob))]

Claims_data[ duplicated(c_insdid)]
Claims_data[ duplicated(c_insdid) | duplicated(c_insdid, fromLast = T)]

Claims_data[ duplicated(pseudocode_1)]
Claims_data[ duplicated(pseudocode_1) | duplicated(pseudocode_1, fromLast = T)]
Claims_data[ duplicated(pseudocode_2)]
Claims_data[ duplicated(pseudocode_2) | duplicated(pseudocode_2, fromLast = T)]

```

### Saving Prepped Data

```{r}
# Creating New Directory from R
dir.create("RData/1. Prepped Data")

saveRDS(Policy_data, "RData/1. Prepped Data/Policy_data.rds")
saveRDS(Claims_data, "RData/1. Prepped Data/Claims_data.rds")

```

## 5) Merging data / Crosschecking Data

```{r}

Policy_data <- readRDS("RData/1. Prepped Data/Policy_data.rds")
Claims_data <- readRDS("RData/1. Prepped Data/Claims_data.rds")

all( Claims_data$c_insdid %in% Policy_data$i_insdid)
all( Claims_data$c_polnum %in% Policy_data$i_polnum)

Policy_data[ , .N]
Claims_data[ , .N]


# Doing Left join onto Policy Data
combined_data <-
  merge(Policy_data, Claims_data, by.x = "i_insdid", by.y = "c_insdid", all.x = T)
combined_data[,.N]

# Checking that all Claims data have a corresponding match
all( Claims_data$c_seqid %in% combined_data$c_seqid)

combined_data[ , table(c_clmct, c_clmsts, useNA = "ifany")]
combined_data[ , table(i_endsts, c_clmsts, useNA = "ifany")]

saveRDS(combined_data[ c_clmsts == "D" & i_endsts == "T"], "RData/0. Misc Data/declined_and_terminated.rds")

# FOr those that are declined, we consider them to be in-force, also to check back later on
combined_data[ c_clmsts == "D" , 
               `:=`( i_endsts = "I",
                     i_terdt  = NA_Date_)]

# Keep columns with "c_" pattern and keep lines with data
Claims_data <- combined_data[ , .SD , .SDcols = patterns( "c_" )]
Claims_data <- Claims_data[ !is.na(c_seqid)]

# Keep columns with "i_" pattern 
Policy_data <- combined_data[ , .SD , .SDcols = patterns( "i_" )]

# Creating Directory
dir.create("RData/2. Data for exposure")

saveRDS(Policy_data, "RData/2. Data for exposure/Policy_data.rds")
saveRDS(Claims_data, "RData/2. Data for exposure/Claims_data.rds")

```
